Statistical Model Selection for Markov Models of Biomolecular Dynamics

### 2014-04-07 paper/paper2.tex
figure updates and fixed the citation issues

\section{Introduction}
Proteins are highly complex molecular machines, and their dynamics are an essential aspect of biomolecular function. These dynamics span a wide range of length scales, timescales and complexity, including folding and aggregation, conformational change between functional native sub-states, ligand binding, and allostery \cite{Dobson2003Protein, Kim2008Real, Austin1975Dynamics, Bahar2007Intrinsic}. Whereas classical experimental probes have often been interpreted in two-state frameworks, ensemble measurements with increasingly high temporal resolution as well as sensitive single molecule probes have uncovered a vast array of complex multi-state kinetics \cite{Cosa2006Evidence, Zhang2011Direct}. But the atomic-resolution characterization of these dynamics is often an enormous challenge -- as molecular probes like F\"{o}rster resonance energy transfer, small-angle x-ray scattering, and nuclear magnetic resonance techniques measure complex projections of the intrinsic structure, generally reporting simultaneously on many degrees of freedom\cite{FRET_PAPER?, Mertens2010Structural, Tzeng2011Protein}.

@@

Two competing sources of error govern the accuracy of an MSM. The first source of error is systematic and is common to all dimensionality reduction methods, in that simplified representations of complex systems discard information. The second source of error is statistical in nature, in that the parameters of an MSM must be estimated from a finite number of stochastic MD trajectories. As we discuss below, these two sources of error compete in a bias-variance tradeoff\cite{bias_variance_paper}. Given a finite size data set, as the number of states increases, the systematic error (the bias) decreases, while the statistical error (the variance) increases.

@@

\section{Background}
Let $\mathbf{x}_t$ be a ergodic time-homogenous Markov process the phase space $\Omega$ with stationary density $\mu(\mathbf{x})$. Consider an ensemble of such processes at time $t$, described by a distribution $p_t(\mathbf{x})$. There exists an operator, termed the propagator, that evolves the a distribution forward in time at discrete intervals, $\tau$.
\begin{equation}
p_{t+\tau}(\mathbf{x}) = \mathcal{Q}^{(\tau)} \circ p_{t}(\mathbf{x})
\end{equation}
The eigenfunctions of the propagator form a basis for $p(\mathbf{x})$, and the time evolution of the ensemble after $n$ applications of the propagator can be decomposed into a sum of terms along each eigenfunction.
\begin{equation}
\label{eq:prop-timescales}
p_{t+n\tau}(\mathbf{x}) = \sum_{i=1}^\infty \exp\left(-\frac{n \cdot \tau}{t_i}\right) \langle \phi_i, p_t \rangle_{\mu^{-1}} \phi_i(\mathbf{x}),
\end{equation}
where $t_i = -\displaystyle\frac{\tau}{\ln \lambda_i}$ are the propagator's relaxation timescales and $\phi_i$ and $\lambda_i$ are the eigenfunctions and eigenvalues of the propagator, and $\langle \cdot, \cdot \rangle_{\mu^{-1}}$ denotes the inner product defined as
\begin{equation}
\langle \phi_i, p_t \rangle_{\mu^{-1}} = \int_\Omega d\mathbf{x}\; \phi_i(\mathbf{x}) p_t(\mathbf{x}) \frac{1}{\mu(\mathbf{x})}.
\end{equation}
For a more detailed discussion of the propagator and its properties, we refer the reader to \citet{Prinz2011Generation}.

For classical MD, $\mathcal{Q}^{(\tau)}$ is uniquely determined by the Hamiltonian and equations of motion\cite{}. However, it is not practical to explicitly construct the propagator from the Hamiltonian for complex systems. A Markov model provides an estimate for the propagator. A $k$-state Markov model is defined on a discrete set of states, $S = \{s_i\}_{i=1}^k$, where $s_i \subseteq \Omega$ and $\bigcup_{i=1}^k s_i = \Omega$. Furthermore, $s_i \cap s_j = \emptyset$ for all $i \neq j$. In words, every point in $\Omega$ is assigned to one (and only one) state in the MSM. Let $\sigma(\mathbf{x})$ be the function that maps a point $\mathbf{x} \in \Omega$ to the index $i$ of the state in $S$ such that $x \in s_i$.

@@

Let $D$ be a set of molecular dynamics trajectories in $\Omega$. To construct an MSM, both $S$ and $\mathbf{T}$ must be determined from the data. The state-space, $S$, can be constructed by clustering the points in $D$ or via a grid-based discretization of $\Omega$. Given $S$, maximum-likelihood estimators for the $O(k^2)$ transition probabilities $\mathbf{T}$ exist\cite{kyle, frank}.

Using this procedure, it has been shown that MSM estimates of the eigenvalues $\lambda_i$ are systematically underestimated in the limit of infinite data, and that the magnitude of this bias goes to zero as $k\rightarrow \infty$. This fact has inspired the development of variational methods for model selection, which optimize the MSM parameters in order to maximize the eigenvalues \cite{feliks}.

Selecting the size of $S$ is an important step in model construction, however there is not currently a widely-accepted method in use. Heuristic criteria, such as manually selecting $k = n_\textnormal{conformation} / 14$ or clustering $D$ such that each conformation is within a pre-specified distance to its cluster center (e.g. 4.5\AA$\;$ or 1.2\AA $\;$ RMSD) have been previously employed\cite{Lane2012, Bowman2012, Beuchamp}. Selection of $\tau$ based convergence of the relaxation timescales (\cref{eq:msm-timescales}), as suggested by \citet{Swope2004Describing} and others is more well established. The rate of this convergence with respect to $\tau$ at fixed $S$ has also been used as a metric for evaluating a model's state decomposition. \hl{cite someone?}

\section{The Bias-Variance Dilemma}
For MSM methods to be valuable to practitioners, they must operate in limited data regimes. Thus, we seek a method for model selection that is sensitive to both the systematic error discussed above, as well as the statistical error associated with estimating the model's parameters. Whereas model selection based on the variational principle described above would tend to increase the number of states without bound, as the magnitude of the eigenvalue bias decreases with the number of states, practical application of this criteria would cause a catastrophic increase in statistical uncertainty in the model. This tradeoff between low-variance (small $k$) but biased estimators, and high-variance (large $k$) but unbiased estimators is a general feature of problems in statistical learning\cite{}. For example, it can be shown that the uncertainty in the posterior distribution of $\mathbf{T}$ given $S$ and $D$ approaches infinity as $k$ approaches infinity.

@@

\section{Likelihood of a Markov State Model}
Bayes' rule provides a foundation for model selection, by establishing the proportionality of the probability of a model given the data to the probability of the data given the model. Without a strong prior, model selection is reduced to the search for the model that maximizes the likelihood of the data. 
With $\{\mathbf{x}_t\}_{t=1}^N$ an observed trajectory of length $N$, the likelihood can be written as:
\begin{equation}
P\big(\{\mathbf{x}_t\}_{t=1}^N\;\big|\; \{S, T\}\big)d\mathbf{x}^N = \prod_{t=1}^{N-1} T_{\sigma(\mathbf{x}_t), \sigma(\mathbf{x}_{t+1})} \cdot \prod_{t=1}^{N} P(\mathbf{x}_t | \sigma(\mathbf{x}_t)) \cdot d\mathbf{x}^N,
\label{eq:like}
\end{equation} 
where $P(\mathbf{x}_t | \sigma(\mathbf{x}_t))$ is the probability of sampling a conformation $\mathbf{x}_t$ given that the process is in the state $\sigma(\mathbf{x}_t)$, referred to as the emission distribution of state $\sigma(\mathbf{x}_t)$. There is some flexibility in the choice of emission distribution, though \cref{eq:like} is valid only when the emission distributions satisfy
\begin{equation}
P(\mathbf{x} | i) = 0 \;\;\;\; \forall \; \mathbf{x} \textnormal{ s.t. } \sigma(\mathbf{x}) \neq i.
\end{equation}
Given the coarse-graining of $p(\mathbf{x})$ in \cref{eq:coarse-grain-p}, the most natural choice emission distributions are normalized indicator functions
\begin{equation}
P(\mathbf{x} \;|\; i) = \frac{1}{V_i} \mathbf{1}_{s_i}(\mathbf{x}).
\end{equation}
An alternative emission distribution was proposed by \citet{Kellogg2012Evaluation} Instead of having support on $\Omega$, this distribution is supported on the training set $\{\mathbf{x}_t\}_{t=1}^N$. It is given by
\begin{equation}
P(\mathbf{x} \;|\; i) = \big| \{ \mathbf{x}_k : \sigma(\mathbf{x}_k) = \sigma(\mathbf{x}) \} \big|^{-1} \sum_{t=1}^N \delta(\mathbf{x}-\mathbf{x}_t),
\end{equation} where $| \cdot |$ denotes the cardinality of a set and $\delta(\cdot)$ is the Dirac delta function. As this is a discriminative model, it is unable to generalize and assign probability to new data. In certain circumstances, this is an undesirable property. For example, protocols that involve fitting and validating models on separate data sets (e.g. cross-validation) are impossible when the statistical model lacks the capacity to describe new data.

@@

Next, we reanalyzed two ultra-long 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain\cite{Liu2008Experimental}, provided courtesy of D.E. Shaw Research \cite{Shaw2010Atomic} (Amber ff99SB-ILDN force field,\cite{amber} TIP3P water model\cite{tip3p}). We projected the trajectories into a four dimensional space using time-structure based independent components analysis (tICA) \cite{Schwantes2013Improvements, Perez2013Identification} on the first trajectory. To calculate the tICs, each conformation was represented as a vector of distances between all pairs of residues separated by at least three amino acids. The distance $d(A, B)$ between residues $A$ and $B$ was determined as follows. Let $\{a_i\}_{i=1}^{n_a}$ and $\{b_i\}_{i=1}^{n_b}$ be the cartesian coordinates of the heavy (non-hydrogen) atoms in $A$ and $B$. Then we define
$$
d(A,B) \equiv \min_{i, j} ||a_i - b_j||_2,
$$ where $|| \cdot ||_2$ denotes the $\ell_2$ norm. The tICs were computed using a correlation lag time of 200 ns. We also performed Principal Components Analysis (PCA) on the same residue-residue distance representation and built MSMs using the top three PCs. In each projection, $k$-centers was used with the Euclidean distance metric.

@@

Leaving out a portion of a molecular dynamics data set from the training to serve as a test set is costly proposition for the applications of MSMs to the study of complex system. For large bimolecular simulations characterized by long intrinsic timescales, sampling the potential exhaustively is a significant challenge\cite{Lane2012}, making it difficult to afford discarding half of the dataset during the fitting of an MSM. For this reason, we also computed two penalized likelihood model selection criteria, the AIC and BIC, which augment the training set log-likelihood with an explicit penalty on model complexity to avoid over fitting. Applied to our simulations of the M\"{u}ller potential, both the AIC and BIC penalize model complexity more strongly than the test set log-likelihood.

@@

\caption{MSMs for the Fip35WW domain built with tICA were evaluated based on four model selection criteria. The log-likelihood was evaluated on the training set (black) and a separate test set (blue), and compared with the AIC (purple) and BIC (red) penalized log-likelihoods, computed on the training set. Dashed lines are used to indicate criterion-minimizing models. The models' ten slowest relaxation timescales (shown in the right panel) converge rapidly. \hl{right panel should be linear scale instead, for consistency with the PCA}}
\end{figure}

@@

\begin{figure}
\centering
\label{fig:ww-pca}
\includegraphics[width=6in]{figs/ww_pca/ww_pca.pdf}
\caption{The four selection criteria were used to compare MSMs for the Fip35WW domain built with PCA. The longest relaxation timescale (right panel) exhibits a linear increase up to 25,000 states. The AIC (blue), BIC (red) and test set log-likelihood each recommend a model with 200 states.} 
\end{figure}

>>>>

\section{Introduction}
Proteins are highly complex molecular machines, and their dynamics are an essential aspect of biomolecular function. These dynamics span a wide range of length scales, timescales and complexity, including folding and aggregation, conformational change between functional native sub-states, ligand binding, and allostery \cite{Dobson2003Protein, Kim2008Real, Austin1975Dynamics, Bahar2007Intrinsic}. Whereas classical experimental probes have often been interpreted in two-state frameworks, ensemble measurements with increasingly high temporal resolution as well as sensitive single molecule probes have uncovered a vast array of complex multi-state kinetics \cite{Cosa2006Evidence, Zhang2011Direct}. But the atomic-resolution characterization of these dynamics is often an enormous challenge -- as molecular probes like F\"{o}rster resonance energy transfer, small-angle x-ray scattering, and nuclear magnetic resonance techniques measure complex projections of the intrinsic structure, generally reporting simultaneously on many degrees of freedom\cite{Lipman2003Single, Mertens2010Structural, Tzeng2011Protein}.

@@

Two competing sources of error govern the accuracy of an MSM. The first source of error is systematic and is common to all dimensionality reduction methods, in that simplified representations of complex systems discard information. The second source of error is statistical in nature, in that the parameters of an MSM must be estimated from a finite number of stochastic MD trajectories. As we discuss below, these two sources of error compete in a bias-variance tradeoff\cite{Sammut2010Encyclopedia}. Given a finite size data set, as the number of states increases, the systematic error (the bias) decreases, while the statistical error (the variance) increases.

@@

\section{Background}
Let $\mathbf{x}_t$ be a ergodic time-homogenous Markov process the phase space $\Omega$ with stationary density $\mu(\mathbf{x})$. Consider an ensemble of such processes at time $t$, described by a distribution $p_t(\mathbf{x})$. There exists an operator, termed the propagator, that evolves the a distribution forward in time at discrete intervals, $\tau$.
\begin{equation}
p_{t+\tau}(\mathbf{x}) = \mathcal{Q}^{(\tau)} \circ p_{t}(\mathbf{x})
\end{equation}
The eigenfunctions of the propagator form a basis for $p(\mathbf{x})$, and the time evolution of the ensemble after $n$ applications of the propagator can be decomposed into a sum of terms along each eigenfunction.
\begin{equation}
\label{eq:prop-timescales}
p_{t+n\tau}(\mathbf{x}) = \sum_{i=1}^\infty \exp\left(-\frac{n \cdot \tau}{t_i}\right) \langle \phi_i, p_t \rangle_{\mu^{-1}} \phi_i(\mathbf{x}),
\end{equation}
where $t_i = -\displaystyle\frac{\tau}{\ln \lambda_i}$ are the propagator's relaxation timescales and $\phi_i$ and $\lambda_i$ are the eigenfunctions and eigenvalues of the propagator, and $\langle \cdot, \cdot \rangle_{\mu^{-1}}$ denotes the inner product defined as
\begin{equation}
\langle \phi_i, p_t \rangle_{\mu^{-1}} = \int_\Omega d\mathbf{x}\; \phi_i(\mathbf{x}) p_t(\mathbf{x}) \frac{1}{\mu(\mathbf{x})}.
\end{equation}
For a more detailed discussion of the propagator and its properties, we refer the reader to \citet{Prinz2011Markov}.

For classical MD, $\mathcal{Q}^{(\tau)}$ is determined by the Hamiltonian and equations of motion\cite{Schutte2001Transfer}. However, it is not practical to explicitly construct the propagator from the Hamiltonian for complex systems. A Markov model provides an estimate for the propagator. A $k$-state Markov model is defined on a discrete set of states, $S = \{s_i\}_{i=1}^k$, where $s_i \subseteq \Omega$ and $\bigcup_{i=1}^k s_i = \Omega$. Furthermore, $s_i \cap s_j = \emptyset$ for all $i \neq j$. In words, every point in $\Omega$ is assigned to one (and only one) state in the MSM. Let $\sigma(\mathbf{x})$ be the function that maps a point $\mathbf{x} \in \Omega$ to the index $i$ of the state in $S$ such that $x \in s_i$.

@@

Let $D$ be a set of molecular dynamics trajectories in $\Omega$. To construct an MSM, both $S$ and $\mathbf{T}$ must be determined from the data. The state-space, $S$, can be constructed by clustering the points in $D$ or via a grid-based discretization of $\Omega$. Given $S$, maximum-likelihood estimators for the $O(k^2)$ elements of a reversible transition probability matrix $\mathbf{T}$ exist\cite{Beauchamp2011Msmbuilder2, Prinz2011Markov}.

Using this procedure, it has been shown that MSM estimates of the eigenvalues $\lambda_i$ are systematically underestimated in the limit of infinite data, and that the magnitude of this bias goes to zero as $k\rightarrow \infty$. This fact has inspired the development of variational methods for model selection, which optimize the MSM parameters in order to maximize the eigenvalues \cite{Noe2013Variational}.

Selecting the size of $S$ is an important step in model construction, however there is not currently a widely-accepted method in use. Heuristic criteria, such as manually selecting $k = n_\textnormal{conformation} / 14$ or clustering $D$ such that each conformation is within a pre-specified distance to its cluster center (e.g. 4.5\AA$\;$ or 1.2\AA $\;$ RMSD) have been previously employed\cite{Lane2011Markov, Bowman2012Equilibrium, Beauchamp2012Simple}. Selection of $\tau$ based convergence of the relaxation timescales (\cref{eq:msm-timescales}), as suggested by \citet{Swope2004Describing} and others is more well established. The rate of this convergence with respect to $\tau$ at fixed $S$ has also been used as a metric for evaluating a model's state decomposition \cite{someone}.

\section{The Bias-Variance Dilemma}
For MSM methods to be valuable to practitioners, they must operate in limited data regimes. Thus, we seek a method for model selection that is sensitive to both the systematic error discussed above, as well as the statistical error associated with estimating the model's parameters. Whereas model selection based on the variational principle described above would tend to increase the number of states without bound, as the magnitude of the eigenvalue bias decreases with the number of states, practical application of this criteria would cause a catastrophic increase in statistical uncertainty in the model. This tradeoff between low-variance (small $k$) but biased estimators, and high-variance (large $k$) but unbiased estimators is a general feature of problems in statistical learning\cite{Sammut2010Encyclopedia}. For example, it can be shown that the uncertainty in the posterior distribution of $\mathbf{T}$ given $S$ and $D$ approaches infinity as $k$ approaches infinity.

@@

\section{Likelihood of a Markov State Model}
Bayes' rule provides a foundation for model selection, by establishing the proportionality of the probability of a model given the data to the probability of the data given the model. Without a strong prior, model selection is reduced to the search for the model that maximizes the likelihood of the data. 
With $\{\mathbf{x}_t\}_{t=1}^N$ an observed trajectory of length $N$, the likelihood can be written as:
\begin{equation}
P\big(\{\mathbf{x}_t\}_{t=1}^N\;\big|\; \{S, T\}\big)d\mathbf{x}^N = \pi_{\sigma(\mathbf{x}_1)}\prod_{t=1}^{N-1} T_{\sigma(\mathbf{x}_t), \sigma(\mathbf{x}_{t+1})} \cdot \prod_{t=1}^{N} P(\mathbf{x}_t | \sigma(\mathbf{x}_t)) \cdot d\mathbf{x}^N,
\label{eq:like}
\end{equation} 
where $\pi_i$ is the stationary probability of state $i$ according to $\mathbf{T}$ and $P(\mathbf{x}_t | \sigma(\mathbf{x}_t))$ is the probability of sampling a conformation $\mathbf{x}_t$ given that the process is in the state $\sigma(\mathbf{x}_t)$, referred to as the emission distribution of state $\sigma(\mathbf{x}_t)$. There is some flexibility in the choice of emission distribution, though \cref{eq:like} is valid only when the emission distributions satisfy
\begin{equation}
P(\mathbf{x} | i) = 0 \;\;\;\; \forall \; \mathbf{x} \textnormal{ s.t. } \sigma(\mathbf{x}) \neq i.
\end{equation}
Given the coarse-graining of $p(\mathbf{x})$ in \cref{eq:coarse-grain-p}, the most natural choice emission distributions are normalized indicator functions
\begin{equation}
P(\mathbf{x} \;|\; i) = \frac{1}{V_i} \mathbf{1}_{s_i}(\mathbf{x}).
\end{equation}
An alternative emission distribution was proposed by \citet{Kellogg2012Evaluation} Instead of having support on $\Omega$, this distribution is supported on the training set $\{\mathbf{x}_t\}_{t=1}^N$. It is given by
\begin{equation}
P(\mathbf{x} \;|\; i) = \big| \{ \mathbf{x}_k : \sigma(\mathbf{x}_k) = \sigma(\mathbf{x}) \} \big|^{-1} \sum_{t=1}^N \delta(\mathbf{x}-\mathbf{x}_t),
\end{equation} where $| \cdot |$ denotes the cardinality of a set and $\delta(\cdot)$ is the Dirac delta function. As this is a discriminative model, it is unable to generalize and assign probability to new data. In certain circumstances, this is an undesirable property. For example, protocols that involve fitting and validating models on separate data sets (e.g. cross-validation) are impossible when the statistical model lacks the capacity to describe new data.

@@

Next, we reanalyzed two ultra-long 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain\cite{Liu2008Experimental}, provided courtesy of D.E. Shaw Research \cite{Shaw2010Atomic} (Amber ff99SB-ILDN force field,\cite{Lindorff2010Improved} TIP3P water model\cite{Jorgensen1983Comparison}). We projected the trajectories into a four dimensional space using time-structure based independent components analysis (tICA) \cite{Schwantes2013Improvements, Perez2013Identification} on the first trajectory. To calculate the tICs, each conformation was represented as a vector of distances between all pairs of residues separated by at least three amino acids. The distance $d(A, B)$ between residues $A$ and $B$ was determined as follows. Let $\{a_i\}_{i=1}^{n_a}$ and $\{b_i\}_{i=1}^{n_b}$ be the cartesian coordinates of the heavy (non-hydrogen) atoms in $A$ and $B$. Then we define
$$
d(A,B) \equiv \min_{i, j} ||a_i - b_j||_2,
$$ where $|| \cdot ||_2$ denotes the $\ell_2$ norm. The tICs were computed using a correlation lag time of 200 ns. We also performed Principal Components Analysis (PCA) on the same residue-residue distance representation and built MSMs using the top three PCs. In each projection, $k$-centers was used with the Euclidean distance metric.

@@

Leaving out a portion of a molecular dynamics data set from the training to serve as a test set is costly proposition for the applications of MSMs to the study of complex system. For large bimolecular simulations characterized by long intrinsic timescales, sampling the potential exhaustively is a significant challenge\cite{Lane2013Milliseconds}, making it difficult to afford discarding half of the dataset during the fitting of an MSM. For this reason, we also computed two penalized likelihood model selection criteria, the AIC and BIC, which augment the training set log-likelihood with an explicit penalty on model complexity to avoid over fitting. Applied to our simulations of the M\"{u}ller potential, both the AIC and BIC penalize model complexity more strongly than the test set log-likelihood.

@@

\caption{MSMs for the Fip35WW domain built with tICA were evaluated based on four model selection criteria. The log-likelihood was evaluated on the training set (black) and a separate test set (blue), and compared with the AIC (purple) and BIC (red) penalized log-likelihoods, computed on the training set. Dashed lines are used to indicate criterion-minimizing models. The models' ten slowest relaxation timescales (shown in the right panel) converge rapidly. }
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=6in]{figs/ww_pca/ww_pca.pdf}
\caption{The four selection criteria were used to compare MSMs for the Fip35WW domain built with PCA. The longest relaxation timescale (right panel) exhibits a linear increase up to 25,000 states. The AIC (blue), BIC (red) and test set log-likelihood each recommend a model with 200 states.\label{fig:ww-pca}} 
\end{figure}
